---
title: "STAT3333_Final_projectrmd"
author: "Marali Benitez, Kelly Thackeray, Lance John"
date: "2025-11-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(broom)
library(infer)   # for permutation/bootstrap tests
```

##                              Technology Useage & Stress Wellness

This project analyzes the relationship between technology usage (daily screen time, phone usage) and wellness outcomes (stress level, mental health score, sleep quality).
We investigate differences across age groups, genders, and location types.
Research Questions:
• 	Do stress levels differ significantly by gender or age group?
• 	Is screen time correlated with mental health scores?
• 	Are stress and physical activity independent?
• 	Can regression models predict stress from tech use?
Relevance: Understanding how tech use impacts wellness can inform healthier digital habits and public health strategies.

```{r}
# Load the data
data <- read.csv("Tech_Use_Stress_Wellness.csv")
# Exploratory Data Analysis
#View(data)
#str(data)
summary(data)
ggplot(data, aes(x = daily_screen_time_hours)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Daily Screen Time")
```

```{r}

# Daily screen time by gender
ggplot(data, aes(x = daily_screen_time_hours, fill = gender)) + labs(title = "Daily Screen Time per Gender", x = "Daily Screen Time (hours)", y = "Count per Gender" ) + geom_histogram() + theme_minimal()

# Correlation between daily screen time and stress level
ggplot(data, aes(x = daily_screen_time_hours, y = stress_level, color = gender)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Screen Time vs Stress by Gender",
       x = "Daily Screen Time (hours)",
       y = "Stress Level") + theme_minimal()
# Wellness
ggplot(data, aes(x = daily_screen_time_hours, y = mental_health_score)) +
  geom_point(alpha = 0.6, color = "steelblue", size = 2) +   # softer points
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linetype = "dashed", size = 1.2) +
  labs(
    title = "Daily Screen Time vs Mental Health Score",
    x = "Daily Screen Time (hours)",
    y = "Mental Health Score") + theme_minimal() 

# Summary statistics
data %>%
  group_by(gender) %>%
  summarise(mean_stress = mean(stress_level, na.rm = TRUE),
            mean_wellness = mean(mental_health_score, na.rm = TRUE))
data %>%
  group_by(location_type) %>%
  summarise(mean_stress = mean(stress_level, na.rm = TRUE),
            mean_wellness = mean(mental_health_score, na.rm = TRUE))
data %>%
  group_by(age) %>%
  summarise(mean_stress = mean(stress_level, na.rm = TRUE),
            mean_wellness = mean(mental_health_score, na.rm = TRUE))
# Mean daily screen time by location type
data %>%
  group_by(location_type) %>%
  summarise(mean_screen_time = mean(daily_screen_time_hours, na.rm = TRUE))
```
```{r}
# Compare CI for stress levels across gender/age groups.
# Bootstrap CI for mean stress level
boot_dist <- data %>%
  specify(response = stress_level) %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "mean")

boot_ci <- boot_dist %>%
  get_ci(level = 0.95, type = "percentile")
boot_ci

```

```{r}
# Hypothesis test
# Example: Is mean stress > 50?
t.test(data$stress_level, mu = 50)

# Compare stress by gender standard t.test insufficient due to 3 factors - use pairwise t.test
pairwise.t.test(data$stress_level, data$gender, p.adjust.method = "bonferroni")
```

```{r}
# Correlation between activity and stress
cor.test(data$physical_activity, data$stress_level)

# Linear regression: stress ~ screen time
lm_model <- lm(stress_level ~ daily_screen_time_hours, data = data)
summary(lm_model)

# Plot regression
ggplot(data, aes(x = daily_screen_time_hours, y = stress_level)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linetype = "dashed") +
  labs(title = "Screen Time vs Stress Level")+ theme_minimal()

```

```{r}
# ANOVA: stress by age group
data$age_group <- cut(
  data$age,
  breaks = c(0, 25, 40, 60, 80, 100),   # adjust ranges as needed
  labels = c("≤25", "26-40", "41-60", "61-80", "81+"),
  right = TRUE
)
anova_result <- aov(stress_level ~ age_group, data = data)
summary(anova_result)
TukeyHSD(anova_result)
ggplot(data, aes(x = age_group, y = stress_level)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Stress Level by Age Group")
```

```{r}
# Chi-square test for independence between stress level (high/low) and physical activity (high/low)
chisq.test(table(data$mood_rating, data$sleep_quality))
```

Conclusion
• 	Stress levels vary across demographic groups.
• 	Screen time shows a measurable correlation with mental health scores.
• 	ANOVA suggests differences in stress across age groups.
• 	Independence tests reveal relationships between categorical wellness factors.

References
• 	Kaggle Dataset: Tech Use & Stress Wellness
• 	Chihara, L., & Hesterberg, T. (2019). Mathematical Statistics with Resampling and R.